---
title: "ANOVA"
---

We will use a built-in dataset of chick weights fed on different types of feed. The dataset can be accessed using the built-in variable `chickwts`.  First, lets plot some boxplots using `qplot()`.

```{r boxplot, fig.width=10}
library(ggplot2)
qplot(x=feed, y=weight, data=chickwts, geom="boxplot", fill=feed) + 
  theme_bw(20)
```

From the boxplots, it sure looks like there **might** be differences in weight between the different feed types.  To be sure, we should do an ANOVA on the data. You can do an ANOVA in R using the linear model function `lm()`, like so. 

## using `lm()`

```{r ANOVA}
myANOVA <- lm(weight ~ feed, data=chickwts)
```


Nothing apears to happen, because we have saved the output to a variable: `myANOVA`.  Now lets output the results as a standard ANOVA table. 

## Standard ANOVA table

```{r anovaTable}
anova(myANOVA)
```

This informs us that the categorical variable `feed` has a highly significant effect on the weight variable `weight`, but it doesn't tell us anything more specific about the differences between differnt levels of the `feed` variable.  We can get a little more information with the `summary()` function.

## Summarizing linear models with `summary()`

```{r summaryANOVA}
summary(myANOVA)
```