---
title: "Getting data in and out"

---
## Input

Getting data into R can be somewhat frustrating for new users.  This is mainly because we humans (the author especially) tend to be sloppy when collecting data in a spreadsheet.  R has rigid expectations.  It expects each column of a text file (usually exported from an Excel sheet) to have a single type of data (text, numeric, factor), plus an optional header row. 

The primary function for reading data into R is the `read.table()` function.  This function reads plain text files (.txt or .csv) and creates a dataframe in R from the resulting data.  There are a variety of arguments to `read.data()`, the most important of which are:

*  `file` = The full path to the file on your computer as a text string. (Use the forward slash `/`, even on Windows.)
*  `sep` = the character that separates columns in your text file.  The default is `" "`, which kind of sucks because most of our files will be separated by either `\t` or `,`.
*  `header` = Whether or not there is a header row in your text file.  Defaults to `FALSE`, but usually we need it to be `TRUE`.

#### <span class="octicon octicon-puzzle"></span> Challenge

Read in the file called femur_lengths.txt. Save to an object called `femora`.

## Output

Usually, I would recommend not writing out modified data frames to text files.  Far better to have a single input file, and to do all necessary manipulations in your saved R script file. But, if you have huge datasets or need to use another statistical program, it is sometimes necesary to save a dataframe out as a text file.  You can do this with `write.table()`

*  `x` = The name of the data frame to be saved. 
*  `file` = Path to the output file.  You can't use `file.choose()` though.
*  `quote` = Do you want quotation marks around strings? Defaults to `TRUE`
*  `sep` = Same as for `read.table()`
*  `row.names` = Do you want row names?  Defaults to `TRUE`, but usually you will set to `FALSE`/

#### <span class="octicon octicon-puzzle"></span> Challenge

Add a new column of log transformed lengths to the femur length dataset and then write out the new file to the Desktop.

```{r include=FALSE}
library(RCurl)
theFile <- getURL("https://dl.dropboxusercontent.com/u/1648032/datasets_for_R_course/femur_lengths.txt")
femora <- read.table(text=theFile, header=TRUE, sep = "\t")
```

```{r include=FALSE}
femora$family <- c(rep("Bovidae", 10), rep("Muridae", 12), rep("Hominidae", 18))
write.table(femora, "~/Desktop/femur_lengths_family.txt", quote=FALSE, sep="\t", row.names=FALSE)
```